Predicting Participants' Vote Choice Between Conservative and Liberty in the 2025 Election By Binary Linear Regression Model

### 2024-04-08 paper/paper.qmd
Add files via upload

---
title: "Predicting the Outcome of the 2025 Election in Canada by a Regression Analysis"
title-block-banner: true
abstract: "ojjojojojljk,mkmk."
thanks: "Code and data supporting this analysis are available at: https://github.com/CSCmaster/Final-Project"
author: "Jingyi Shen"
date: today
date-format: long
format: pdf
toc: true
toc-depth: 2
number-sections: true
bibliography: references.bib
link-citations: true
---

```{r setup, include=FALSE, message=FALSE}
## Import necessary libraries
library(ggplot2)
library(dplyr)
library(tidyverse)
library(here)
library(readr)
library(knitr)
library(kableExtra)
library(rstanarm)
library(arrow)
```

@@

### Data Summary Measures 
Table 1: Summary varibales in Census and Survey Data
```{r echo=FALSE}
numeric_summary_df <- data.frame(
  Variable = c("Age in Census Data", "Age in Survey Data", 
               "Liberal Votes in Survey", "Conservative Votes in Survey"),
  Minimum = c(15.0, 18.0, 0.000, 0.000),
  Q1 = c(37.0, 36.0, 0.000, 0.000),
  Median = c(54.0, 53.0, 0.000, 0.000),
  Average = c(52.18, 51.3, 0.267, 0.249),
  Q3 = c(67.0, 66.0, 1.000, 0.000),
  Maximum = c(80.0, 97.0, 1.000, 1.000)
)

knitr::kable(numeric_summary_df, caption = "Summary of Numerical Variables Across Census and Survey Data")

The statistical summary of the census and survey data reveals age distributions with the census data capturing a narrower age range (15 to 80 years, with a median of 54) compared to the broader age span in the survey data (18 to 97 years, with a median of 53). The average ages for the census and survey populations are 52.18 and 51.3 years, respectively. Voting data from the survey indicate that a quarter of the respondents are inclined towards the Liberal Party, as shown by an average of 0.267, while a slightly lower average of 0.249 suggests Conservative support; both parties have 50% of the population not voting for them, with a median of 0. The upper bounds for Liberal and Conservative votes are 1, meaning some respondents indicated a vote for these parties, while the third quartile for Conservative votes indicates less overall support compared to the Liberal Party.

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
df_long <- pivot_longer(na.omit(survey_data_new), cols = starts_with("vote_"), names_to = "Party", values_to = "Vote")

# Plot a grouped bar plot
ggplot(df_long, aes(x = Party, fill = as.factor(Vote))) +
  geom_bar(position = "dodge", stat = "count") +
  labs(title = "Plot.6 Comparison of Voting Preferences", x = "Party", y = "Count") +
  scale_fill_manual(values = c("red", "blue"), name = "Party Choice", labels = c("Did Not Vote", "Voted")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

@@

$\beta_0$ also known as the model's intercept, is the base probability of voting in a certain way when all explanatory variables are set to their reference levels.

$\beta_1$ is for every one-unit increase in age, it quantifies the corresponding variation in the log odds of voting preference.

$\beta_2$ is the influence on the log odds of preferring a particular voting option when the person is identified as male, with $x_{male}$ being assigned a value of 1 for males and 0 for females.

$\beta_3$ to is the coefficients detail the adjustments in the log odds of the voting choice associated with residency in different provinces.

$x_{BC}$, for example, is set to 1 if the person resides in British Columbia, with similar binary indicators for other provinces from $x_{MB}$ through $x_{SK}$.

$\beta_{4}$ captures the shift in log odds of the voting decision when the individual professes a religious belief, with $x_{religion}$ set to 1 for believers and 0 for non-believers.

$\beta_{5}$ is the modification in the log odds of a particular voting outcome for individuals identified as part of the middle class, where $x_{mid}$ is 1 for middle-class individuals and 0 otherwise.

$\beta_{5}$ measures the alteration in the log odds of voting behavior for individuals who fall within the upper middle class to wealthy category, with $x_{upper}$ indicating this socio-economic status by being set to 1, and 0 for others.

@@

# Appendix

>>>>

---
title: "Predicting Participants' Vote Choice Between Conservative and Liberty in the 2025 Election By Binary Linear Regression Model"
title-block-banner: true
abstract: "ojjojojojljk,mkmk."
thanks: "Code and data supporting this analysis are available at: https://github.com/CSCmaster/Final-Project"
author: "Jingyi Shen"
date: today
date-format: long
format: pdf
toc: true
toc-depth: 2
number-sections: true
bibliography: references.bib
link-citations: true
---

```{r setup, include=FALSE, message=FALSE}
## Import necessary libraries
library(ggplot2)
library(dplyr)
library(tidyverse)
library(here)
library(readr)
library(knitr)
library(kableExtra)
library(rstanarm)
library(arrow)
library(tidybayes)
```

@@

### Data Summary Measures 
```{r}
#| label: tbl-data-summary
#| tbl-cap: Summary of Numerical Variables Across Census and Survey Data
#| echo: false
numeric_summary_df <- data.frame(
  Variable = c("Age in Census Data", "Age in Survey Data", 
               "Liberal Votes in Survey", "Conservative Votes in Survey"),
  Minimum = c(15.0, 18.0, 0.000, 0.000),
  Q1 = c(37.0, 36.0, 0.000, 0.000),
  Median = c(54.0, 53.0, 0.000, 0.000),
  Average = c(52.18, 51.3, 0.267, 0.249),
  Q3 = c(67.0, 66.0, 1.000, 0.000),
  Maximum = c(80.0, 97.0, 1.000, 1.000)
)

knitr::kable(numeric_summary_df)
```

@tbl-data-summary summaries the census and survey data reveals age distributions with the census data capturing a narrower age range (15 to 80 years, with a median of 54) compared to the broader age span in the survey data (18 to 97 years, with a median of 53). The average ages for the census and survey populations are 52.18 and 51.3 years, respectively. Voting data from the survey indicate that a quarter of the respondents are inclined towards the Liberal Party, as shown by an average of 0.267, while a slightly lower average of 0.249 suggests Conservative support; both parties have 50% of the population not voting for them, with a median of 0. The upper bounds for Liberal and Conservative votes are 1, meaning some respondents indicated a vote for these parties, while the third quartile for Conservative votes indicates less overall support compared to the Liberal Party.

@@

```{r, echo=FALSE, message=FALSE, warning=FALSE}
df_long <- pivot_longer(na.omit(cleaned_survey_data), cols = starts_with("vote_"), names_to = "Party", values_to = "Vote")

ggplot(df_long, aes(x = Party, fill = as.factor(Vote))) +
  geom_bar(position = "dodge", stat = "count") +
  labs(title = "Plot.6 Comparison of Voting Preferences", x = "Party", y = "Count") +
  scale_fill_manual(values = c("red", "blue"), name = "Party Choice", labels = c("Did Not Vote", "Voted")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

@@

We define our Bayesian logistic regression model for the probability of an individual voting for the Liberal/Conservative party as follows:

- $y_i$: Binary outcome of voting for the Liberal/Conservative party (1) or not (0)
- $\pi_i$: Probability of voting for the Liberal/Conservative party

\begin{align}
y_i | \pi_i &\sim \text{Bernoulli}(\pi_i) \\
\text{logit}(\pi_i) &= \beta_0 + \alpha^{age}_{a[i]} + \alpha^{gender}_{g[i]} + \alpha^{province}_{p[i]} + \alpha^{religion}_{r[i]} + \alpha^{income}_{m[i]} + \alpha^{education}_{e[i]} \\
\beta_0 &\sim \text{Normal}(0, 2.5) \\
\alpha_{a}^{age} &\sim \text{Normal}(0, \sigma^{2}_{age}) \quad \text{for } a \text{ in } 1, 2, \ldots, A \\
\alpha_{g}^{gender} &\sim \text{Normal}(0, \sigma^{2}_{gender}) \quad \text{for } g \text{ in } Male,Female \\
\alpha_{p}^{province} &\sim \text{Normal}(0, \sigma^{2}_{province}) \quad \text{for } p \text{ in } BC, AB, MB, NB, NL, NS, ON, PE, QC, SK \\
\alpha_{r}^{religion} &\sim \text{Normal}(0, \sigma^{2}_{religion}) \quad \text{for } r \text{ in } YES,NO \\
\alpha_{m}^{income} &\sim \text{Normal}(0, \sigma^{2}_{income}) \quad \text{for } m \text{ in } Lower Class, Middle Class, Upper Class\\
\alpha_{e}^{education} &\sim \text{Normal}(0, \sigma^{2}_{education}) \quad \text{for } e \text{ in } Low Education, Some Education, High Education \\
\sigma^{2}_{age} &\sim \text{Exponential}(1) \\
\sigma^{2}_{gender} &\sim \text{Exponential}(1) \\
\sigma^{2}_{province} &\sim \text{Exponential}(1) \\
\sigma^{2}_{religion} &\sim \text{Exponential}(1) \\
\sigma^{2}_{income} &\sim \text{Exponential}(1) \\
\sigma^{2}_{education} &\sim \text{Exponential}(1) 
\end{align}

@@

$\beta_0$(Intercept): This is the base log odds of an individual voting for the political party when all the explanatory variables are at their reference levels.

$\beta_1$(age): This coefficient represents the change in the log odds of voting for the political party for each additional year of age. A positive coefficient indicates that as age increases.

$\beta_2$(gender): This coefficient indicates the difference in the log odds of voting for the political party for males compared to females (the reference category). A positive coefficient suggests that males are more likely to vote this party than females.

$\beta_3$(province): Each of these coefficients (e.g. $\beta_BC$, $\beta_ON$, etc) measures the change in the log odds of voting for the political party for individuals living in each respective province compared to the reference province. Each coefficient reflects the unique effect of residing in that province on voting behavior.

$\beta_4$(religion): This coefficient captures the change in the log odds of voting this political party if an individual professes a religious belief (1) versus no religious belief (0, the reference category).

$\beta_5$(income): This coefficients reflect the differences in the log odds of voting for the political party based on income levels.

$\beta_6$(education): Similar to income, these coefficients represent the effect of education level on the log odds of voting for this political party.

@@

# Appendix {.unnumbered}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

```{r echo=FALSE, warning=FALSE}
pp_check(liberal_model, type = "hist")
```
Histogram of posterior predictive distributions from the liberal model. This plot compares the observed counts of the number of votes for the Liberal party (indicated by the dark line) with the simulated counts from the posterior predictive distribution (shown as histograms). The overlap between the observed counts and the predictive distribution suggests how well the model captures the data. Areas where the observed counts fall outside the bulk of the simulated distributions may indicate model misfit.

```{r echo=FALSE, warning=FALSE}
pp_check(conservative_model, type = "hist")
```

```{r echo=FALSE, warning=FALSE,  message=FALSE, fig.width=8, fig.height=6}
survey_data <- read_csv(here("data", "analysis_data", "cleaned_survey_data.csv"))
prior_covariates <- normal(location = 0, scale = 2.5, autoscale = FALSE)
prior_intercept <- normal(0, 2.5)
pp_plot <- posterior_vs_prior(liberal_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10),
    axis.title.y = element_text(size = 7, face = "bold"),
    axis.title.x = element_text(size = 7, face = "bold"),
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 7)
  ) +
  coord_flip() +
  ggtitle("Comparison of Posterior and Prior Distributions",
          subtitle = "This plot visualizes how posterior distributions of model parameters compare with their specified priors for the liberal model.") +
  xlab("Density") +
  ylab("Parameters")

# Print the plot
print(pp_plot)
```

```{r echo=FALSE, warning=FALSE,  message=FALSE, fig.width=8, fig.height=6}
prior_covariates <- normal(location = 0, scale = 2.5, autoscale = FALSE)
prior_intercept <- normal(0, 2.5)
pp_plot <- posterior_vs_prior(conservative_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10),
    axis.title.y = element_text(size = 7, face = "bold"),
    axis.title.x = element_text(size = 7, face = "bold"),
    legend.text = element_text(size = 5),
    legend.title = element_text(size = 7)
  ) +
  coord_flip() +
  ggtitle("Comparison of Posterior and Prior Distributions",
          subtitle = "This plot visualizes how posterior distributions of model parameters compare with their specified priors for the conservative model.") +
  xlab("Density") +
  ylab("Parameters")

# Print the plot
print(pp_plot)

## Diagnostics
